Modeling the effect of aneuploidy on cancer evolution

### 2023-05-31 EvolutionaryRescue1.tex
reorganize figures and rescue results -> now going to merge figures

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Evolutionary simulation} 
Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.

@@

Evolutionary rescue occurs when mutant cells appear and fixate ($m_t>>1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.

To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells.
Thus, the rescue probability is given by 
\begin{equation}
p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w} ,
\end{equation}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

Now, we apply the approximations for the survival probability $p_w$ from \Cref{survprobwapprox1,survprobwinitial,scenario3}. Therefore, substituting $\Delta_k=\lambda_k-\mu_k$ , the rescue probability is given by
\begin{equation}\label{rescue_prob}
p_{rescue} = \left\{
  \begin{array}{@{}ll@{}}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],\quad\text{if }4\lambda_avp_m>\left(\Delta_a-v\right)^2,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],\quad\text{if }\Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right],\quad\text{if }\Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2.
  \end{array}\right.
\end{equation}

%Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
%\begin{align}\label{survprobresistantaneuploidy}
%p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
%\end{align}

@@

\label{SurvPlot}

@@

\label{SurvPlotLargeN}
\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells. The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations. For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
\label{SurvPlotNData}

@@

\label{P_est}

@@

\label{P_est_large_N}

@@

\label{DeleteriousPlot}

@@

\label{SurvPlotNDataLogisticKComplete}

@@

% diffusion approx instead of branching process
\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
\label{P_est_diffusion}

@@

\begin{appendices}
\renewcommand{\theequation}{\thesection\arabic{equation}}
\counterwithin*{equation}{section}

@@

\section*{Diffusion approximation}\label{AppendixDiffApprox}
An alternative method to obtain the probability of evolutionary rescue is to utilize a Feller diffusion approximation which is governed by two parameters: the growth rate $r$ and the reproductive variance $\sigma$. The two parameters are obtained from the underlying demographic process as the infinitesimal relative change in mean and variance of $n_t$ over and an infinitesimally small time interval $\Delta t$:
\begin{align*}
r&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{E}\left(\Delta n_t|n_t\right)}{\Delta_t n_t},\\
\sigma&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{V}ar\left(\Delta n_t|n_t\right)}{\Delta_t n_t}.
\end{align*}
The rate at which mutants are generated directly from the wildtype is: 
\begin{align}
\theta_1=v\bar{\pi}_f\frac{N}{|r_w|},
\end{align}
where
\begin{align}
\bar{\pi}_f=\int_{0}^\infty\int_{0}^\infty \left(1-\e^{-\frac{2r}{\sigma}}\right) f_r\left(r,\sigma\right)\d r\d\sigma.
\end{align}
Letting $f_r\left(r,\sigma\right)=\delta\left(r-r_m\right)\delta\left(\sigma-\sigma_m\right)$ then:
\begin{align}
\bar{\pi}_f=1-\e^{-\frac{2r_m}{\sigma_m}},
\end{align}
and, as a result, we have:
\begin{align}
\theta_1=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right)\frac{N}{|\Delta_w|}.
\end{align}
The rate at which mutants are generated indirectly from the wildtype through aneuploidy is: 
\begin{align*}
\theta_2&=\frac{uN}{|r_w|}\int_{0}^\infty\int_{0}^\infty \left(1-\pi_f\left(r,q\right)\right) f_r\left(r,\sigma\right) p_1^*\left(r,\sigma\right)\d r\d\sigma\\
&=\frac{uN}{|\Delta_w|}p_1^*\left(r_a,\sigma_a\right)\\
&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|r_a|}{\sigma_a^2}\left(\sqrt{1+\frac{2\sigma_a^2}{r_a^2}u^*}-1\right)\right)\right]\\
&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|\Delta_a|}{(\lambda_a+\mu_a)^2}\left(\sqrt{1+\frac{2(\lambda_a+\mu_a)^2}{\Delta_a^2}u^*}-1\right)\right)\right],
\end{align*}
where
\begin{equation}
u^*=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right).
\end{equation}
The number of rescue mutations has a Poisson distribution with rate $\theta_1+\theta_2$. As a result, the probability of evolutionary rescue is given by:
\begin{equation}\label{FellerApprox}
p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},
\end{equation}
which we plot in Figure \ref{P_est_diffusion} as a function of $\Delta_w$.
\end{appendices}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Stochastic simulations} 
Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.

@@

In our model, \emph{evolutionary rescue} occurs when resistant cells appear and fixate ($m_t \gg 1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.

To estimate the rescue probability $p_{rescue}$, we assume independence between clonal lineages starting from an initial population of $N$ wildtype cells (we check the effect of density-dependent growth on our results below).
Thus, the rescue probability is given by 
\begin{equation}
p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w} ,
\end{equation}
where the approximation $(1-p_w)\approx e^{-p_w}$ assumes that $p_w$ (but not $N p_w$) is small.

Applying the approximations for the survival probability $p_w$ from \cref{survprobwapprox1,survprobwinitial,scenario3} and substituting $\Delta_k=\lambda_k-\mu_k$, we find that the rescue probability can be approximated by
\begin{equation}\label{rescue_prob}
\begin{aligned}
&p_{rescue} \approx \\
  &\begin{cases}
  1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right] ,&
  4\lambda_avp_m>\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right] ,&
   \Delta_a<0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 ,\\
   1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] ,&
   \Delta_a>0\quad\text{and}\quad4\lambda_avp_m<\left(\Delta_a-v\right)^2 .
  \end{cases}
\end{aligned}
\end{equation}

% TODO do we need this?
%Note that if instead \cref{survprobwapprox1} we apply \cref{survprobwapproxcorrected}, we have
%\begin{align}\label{survprobresistantaneuploidy}
%p_{rescue}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}-\frac{uv^2\lambda_a\Delta_m^2}{8\lambda_m^2\Delta_a^3}\right)\right].
%\end{align}
%However, both approximations seem to work very well (\Cref{compare_1st_2nd_order_approx}).

We validate these approximations by comparing them to results of stochastic evolutionary simulations. We find that the approximations work very well (\Cref{rescue_prob_wt_growth_small_N,rescue_prob_wt_growth_large_N,rescue_prob_an_growth_small_N,rescue_prob_an_growth_large_N,rescue_prob_N_nongrow,rescue_prob_N_logistic}).

@@

\label{rescue_prob_wt_growth_small_N}

@@

\label{rescue_prob_wt_growth_large_N}

@@

\label{rescue_prob_an_growth_small_N}

@@

\label{rescue_prob_an_growth_large_N}

@@

\label{rescue_prob_N_nongrow}

@@

\label{rescue_prob_N_logistic}

@@

%\begin{figure}[p]
%% diffusion approx instead of branching process
%% tolerance
% \vspace*{1\baselineskip}
%\includegraphics[width=1\textwidth]{Figures/P_est_diffusion.pdf}
%\caption{Plot of the survival probability of an initial population consisting of $w_0=10^{4}$ wildtype cells as a function of $\Delta_a=\lambda_a-\mu_a$ for various values of $\Delta_w=\lambda_a-\mu_a$. The continuous lines represent the exact result \eqref{survprobw} while the dashed lines represent the Feller diffusion approximation \eqref{FellerApprox}. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/w_0}$ where $p$ is the mean probability of evolutionary rescue.}
%\label{P_est_diffusion}
%\begin{figure}[p]
%\vspace*{1\baselineskip}
%\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
%\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells. The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations. For the simulations we have chosen the following parameters: $\lambda_w=0.14, \lambda_a=0.14,\lambda_m=0.14,\mu_w=0.17,\mu_a=0.135,\mu_m=0.13$. The error bars represent $95\%$ confidence interval of the form $p\pm1.96\sqrt{p\left(1-p\right)/n}$ where $p$ is the mean probability of evolutionary rescue and $n=100$ is the number of simulations.}
%\label{compare_1st_2nd_order_approx}

@@

%\begin{appendices}
%\renewcommand{\theequation}{\thesection\arabic{equation}}
%\counterwithin*{equation}{section}

@@

%\section*{Diffusion approximation}\label{AppendixDiffApprox}
%
%An alternative method to obtain the probability of evolutionary rescue is to utilize a Feller diffusion approximation which is governed by two parameters: the growth rate $r$ and the reproductive variance $\sigma$. The two parameters are obtained from the underlying demographic process as the infinitesimal relative change in mean and variance of $n_t$ over and an infinitesimally small time interval $\Delta t$:
%r&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{E}\left(\Delta n_t|n_t\right)}{\Delta_t n_t},\\
%\sigma&=\lim_{\Delta t\rightarrow0}\frac{\mathbb{V}ar\left(\Delta n_t|n_t\right)}{\Delta_t n_t}.
%The rate at which mutants are generated directly from the wildtype is: 
%\begin{align}
%\theta_1=v\bar{\pi}_f\frac{N}{|r_w|},
%\end{align}
%where
%\begin{align}
%\bar{\pi}_f=\int_{0}^\infty\int_{0}^\infty \left(1-\e^{-\frac{2r}{\sigma}}\right) f_r\left(r,\sigma\right)\d r\d\sigma.
%\end{align}
%Letting $f_r\left(r,\sigma\right)=\delta\left(r-r_m\right)\delta\left(\sigma-\sigma_m\right)$ then:
%\begin{align}
%\bar{\pi}_f=1-\e^{-\frac{2r_m}{\sigma_m}},
%\end{align}
%and, as a result, we have:
%\begin{align}
%\theta_1=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right)\frac{N}{|\Delta_w|}.
%\end{align}
%The rate at which mutants are generated indirectly from the wildtype through aneuploidy is: 
%\theta_2&=\frac{uN}{|r_w|}\int_{0}^\infty\int_{0}^\infty \left(1-\pi_f\left(r,q\right)\right) f_r\left(r,\sigma\right) p_1^*\left(r,\sigma\right)\d r\d\sigma\\
%&=\frac{uN}{|\Delta_w|}p_1^*\left(r_a,\sigma_a\right)\\
%&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|r_a|}{\sigma_a^2}\left(\sqrt{1+\frac{2\sigma_a^2}{r_a^2}u^*}-1\right)\right)\right]\\
%&=\frac{uN}{|\Delta_w|}\left[1-\exp\left(-\frac{|\Delta_a|}{(\lambda_a+\mu_a)^2}\left(\sqrt{1+\frac{2(\lambda_a+\mu_a)^2}{\Delta_a^2}u^*}-1\right)\right)\right],
%where
%\begin{equation}
%u^*=v\left(1-\e^{-\frac{2r_m}{\sigma_m}}\right).
%\end{equation}
%The number of rescue mutations has a Poisson distribution with rate $\theta_1+\theta_2$. As a result, the probability of evolutionary rescue is given by:
%\begin{equation}\label{FellerApprox}
%p_{rescue}=1-\e^{-\left(\theta_1+\theta_2\right)},
%\end{equation}
%which we plot in Figure \ref{P_est_diffusion} as a function of $\Delta_w$.

%\end{appendices}
